Methods for Analyzing Models of\\ Rod-Shaped Bacteria

### 2025-12-05 paper/main.tex
rewrite section with division

In the next step, we extend our previously used time-series with successive images including
cell division events which means that we now have to use the image-space metric (see
section~\ref{section:parameter-estimation}).
A practical problem in extending the time series is that the segmentation algorithm was not able to
properly segment the individual bacteria around the division event which meant that we had to omit
parts of the snapshots, leaving us with only incomplete data.

% Since we have already optimized said system, we can reduce the computational demand by choosing the
% parameters which were obtained by our previous attempts and extend them with the parameters that the
% daughter cells will carry.
We optimize for the same parameters as done before.
In addition to assigning growth rates individually, we also need to consider division lengths at
which the bacteria split and the new individual growth rates of the daughter cells as additional
parameters.
Since our simulation involves the division of 4 cells into 8 new ones, in principle, this leads
to a total of $8$ new growth rates and $4$ division lengths.
% But to decrease the size of the parameter space, we assume that the growth rates of the daughter
% cells are identical between every sibling, leaving us at only $4$ new growth rates and a total of
% $8$ parameters.
% Future approaches should aim to loosen this assumption, allowing to optimize every new growth rate.
To optimize this system, we used a similar optimization approach by again seeding with samples
generated via latin-hypercube~\cite{McKay1979}, then applying the differential evolution
algorithm~\cite{Storn1997} and finally a local minimization with Nelder-Mead~\cite{Gao2010}.
We used the value $0.5$ for the parent penalty, which aims to provide a smooth transition between
not accounting for them and assigning full weighting to this difference.
Figure~\ref{fig:celldiffs-progression-with-division} shows the difference between the cell mask as
given by the data and the cell mask predicted by the best fit of our model.
As expected, the fully-optimized system does not produce differences where the parental penalty is
required.
However, its value is frequently used within the the optimization scheme itself.

We can again calculate likelihood profiles for every optimized parameter.
The results displayed in Figure~\ref{fig:likelihood-profiles-comparison-with-division} show that
some curves exhibit a plateau close to the calculated optimum of the estimated parameter.
This concerns the division lengths and some growth rates of the mother bacteria which both determine
if and when a cell divides.
Such behavior can be attributed to the missing images around the division events of the cells which
were not able to be segmented.
\textbf{TODO check this with results}
It can be argued that this leads to a non-identifiability for the first division length while the
remaining parameters are still identifiable although the effect is also notable.
These results highlight a reassuring finding, namely that the method of profile-likelihood is able
to properly display this lack of data.
It therefore further underlines the accuracy and predictive power of our methods.

The image-space metric is computationally much more expensive compared to the direct method.
There are 4 main steps which are executed within any single optimization iteration that contribute
to the total wall time.
These steps are prediction via the computational model ($15.63(21)\%$), mask adjustment of the data
masks ($14.49(19)\%$), mask rendering for the numerical results ($69.2(1.4)\%$) and finally
comparison of the masks ($0.68(5)\%$) (see section \nameref{section:supplement-performance}).
The second step alters the masks provided by the data for every individual numerical simulation.
This is required since we need to ensure that the colors of the data-masks match the colors of the
masks produced by our computational model.
To this end, we need to construct a family tree of the data-bacteria and compare with the
\textit{in-silico} family tree.
We can see that the generation of masks from the numerical results takes up the majority of the work.
In addition, the step of mask adjustment would not have been required within the position-comparison
algorithm.
Combining these two values, the position-comparison algorithm is $\approx 6.13(8)$ times faster.

>>>>

In the next step, we extended our previously used time-series with successive images including
cell division events which requires us to use the Lineage-Aware Image-Space Metric
(section~\ref{section:parameter-estimation}).
A practical challenge in extending the time series arose during data preparation: the segmentation
algorithm was not able to properly label the individual bacteria for timepoints close to division
events which meant that we had to omit a subset of snapshots, leaving us with incomplete temporal
data.
This problem is common for existing segmentation algorithms and may also be encountered in future
studies.
We therefore explicitly include this gap to assess the fidelity of our approach under these
practical challenges.

% Since we have already optimized said system, we can reduce the computational demand by choosing the
% parameters which were obtained by our previous attempts and extend them with the parameters that the
% daughter cells will carry.
In addition to the previously used parameters, we also need to consider division lengths at
which the bacteria split and the new individual growth rates of the daughter cells as additional
parameters.
Since our simulation involves the division of 4 cells into 8 new ones, this amounts to a total of $8$
new growth rates and $4$ division lengths.
% But to decrease the size of the parameter space, we assume that the growth rates of the daughter
% cells are identical between every sibling, leaving us at only $4$ new growth rates and a total of
% $8$ parameters.
% Future approaches should aim to loosen this assumption, allowing to optimize every new growth rate.
To optimize this system, we used a similar optimization approach by again seeding with samples
generated via Latin-Hypercube~\cite{McKay1979}, then applying the differential evolution
algorithm~\cite{Storn1997} and finally a local minimization with Nelder-Mead~\cite{Gao2010}.
We used the value $p=0.5$ for the parent penalty, which aims to provide a smooth transition between
not accounting for them and assigning full weighting to this difference.
Figure~\ref{fig:celldiffs-progression-with-division} shows the difference between the cell mask as
given by the data and the cell mask predicted by the best fit of our model.
As expected, the fully-optimized system does not produce differences where the parental penalty is
required.
However, its value is frequently used within the the optimization scheme itself.

% We can again calculate likelihood profiles for every optimized parameter.
Since further mathematical analysis is required to assess how to utilize the Lineage-Aware
Image-Space metric for maximum-likelihood approaches, we simply display the change of the cost
function when fixing a single parameter.
While only confidence bands for likelihood profiles as presented in the previous sections could
paint a clearer picture with regards to the identifiability of the parameters, we can still
identify some trends.
Figure~\ref{fig:likelihood-profiles-comparison-with-division} shows that some curves (F,G,I,L)
exhibit a plateau close to the calculated optimum of the estimated parameter.
This concerns the division lengths and some growth rates of the mother bacteria which both determine
if and when a cell divides.
Such behavior can be attributed to the missing images around the division events of the cells which
were not able to be segmented.
This presents a restriction on the lower and upper bounds of the parameter value.
However for some parameters (H,J,K) the lower region of the plots contains a slight gradient,
indicating that our mechanistic model is able to provide some level of information about the missing
datapoints.
It therefore further underlines the accuracy and predictive power of our methods.
\textbf{TODO check this with results}

The image-space metric is computationally much more expensive compared to the direct method.
There are 4 main steps which are executed within any single optimization iteration that contribute
to the total wall time.
Rendering of the masks dominates the workload ($69.2(1.4)\%$), followed by the model prediction
($15.63(21)\%$) and mask adjustment ($14.49(19)\%$) while the actual pixel-wise compariosn is
negligible ($0.68(5)\%$) (see also Section \ref{section:supplement-performance}).
The mask-adjustment step is necessary for every individual numerical simulation to ensure that masks
obtained from the data are color-comparable to the ones produced by our computational model.
To this end, we need to construct a set of family trees (forest) of the data-bacteria and compare
with the \textit{in-silico} family trees.
We can see that the generation of masks from the numerical results takes up the majority of the work.
In addition, the step of mask adjustment would not have been required within the position-comparison
algorithm.
Combining these two values, the position-comparison algorithm is $\approx 6.13(8)$ times faster.
